---
title: "SEPTA COVID Data Visualization"
author: "Andrew Simpson"
date: "5/20/2020"
output: html_document
---
<style type="text/css">
body{ /* Normal  */
  font-family: "Open Sans";
  font-size: 14px;
  }
h1.title {
  font-family: "Montserrat";
  font-size: 36px;
  color: `{r} rphl::get_colors("light_blue")`;
}
h3.subtitle {
  font-family: "Montserrat";
  font-size: 28px;
  color: `rphl::get_colors("light_blue")`;
}
h1 { /* Header 1 */
  font-family: "Montserrat";
  font-size: 28px;
  
}
h2 { /* Header 2 */
  font-family: "Montserrat";
  font-size: 22px;
  
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Montserrat";
  
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="100%")

require(tidyverse) ; require(sf) ; require(scales) ; require(lubridate) ; require(rphl)
require(grid) ; require(gridExtra) ; require(corrplot) ; require(stargazer)

#install.packages("devtools")
#devtools::install_github("CityOfPhiladelphia/rphl")
#extrafont::font_import()
#https://standards.phila.gov/guidelines/design-development/brand-elements/typography/

source("./code/functions.R")

#### Import data &  join as needed ----------------------------------------- ####
# census tracts
tract_sf <- read_sf("./inputs/tl_2019_42_tract/tl_2019_42_tract.shp")

# planning districts
planningdistricts_sf <- read_sf("./inputs/Planning_Districts.geojson") %>% st_transform(crs = 4269)

# center city
ccd_sf <- read_sf("./inputs/CCD_Boundary.geojson") %>% st_transform(crs = 4269) %>% mutate(ccd = TRUE)

# pre-prepared ACS data
acs_spread <- read_csv("./prepped_data/acs_spread.csv") %>%
  mutate(income_bucket = factor(
      income_bucket, 
      ordered = TRUE,
      levels = c("very_low_income", "low_income", "mid_income", "high_income", "very_high_income"))
    )

stops_byweek_ridership_sf <- st_read("./prepped_data/stop_data.geojson") %>% 
  mutate(stop_id = as.numeric(stop_id))

# assign census tracts to each bus stop
stops_tracts <- st_join(stops_byweek_ridership_sf,
                            left = FALSE,
                            tract_sf["GEOID"]) %>%
  mutate(GEOID = as.numeric(GEOID)) 

#### Shape the data for visulization  ------------------------------------- ####

# Get bus network-wide ridership by week
weekly_ridership <- stops_byweek_ridership_sf %>%
  group_by(monday) %>%
  summarise(daily_ridership = sum(daily_ons, na.rm = TRUE))

# create df of census tracts with bus ridership by week & district boundaries (ccd, planning)
tracts_weekly_ridership <- stops_tracts %>%
  group_by(GEOID, monday) %>%
  summarise(ridership = sum(daily_ons)) %>%
  right_join(acs_spread) %>%
  st_join(ccd_sf) %>% 
  st_join(planningdistricts_sf)

# create before/after dataframe - using "2020-03-30" as the cut point
comp <- build_comp_df(tracts_weekly_ridership, cut = c("2020-03-16", "2020-03-30"))

# census tract binned by income w/ pre/post covid ridership
income_tracts <- comp %>%
  group_by(income_bucket) %>%
  summarise(num_tracts = n(),
            pre_covid = sum(pre_covid_ridership),
            post_covid = sum(post_covid_ridership),
            diff = post_covid - pre_covid,
            pct_change = diff / pre_covid) %>%
  filter(is.na(income_bucket) == FALSE)

# planning districts binned by income w/ pre/pose covid ridership
income_planning <- comp %>%
  group_by(DIST_NAME.x) %>%
  summarise(pre_covid = sum(pre_covid_ridership),
            post_covid = sum(post_covid_ridership),
            avg_med_income = mean(med_income, na.rm = TRUE),
            diff = post_covid - pre_covid,
            pct_change = diff / pre_covid)

# df that shows delta of ridership for each planning tract over time 
income_planning_linear <- tracts_weekly_ridership %>%
  group_by(DIST_NAME, monday) %>%
  summarise(daily_ridership = sum(ridership, na.rm = TRUE)) %>%
  arrange(monday) %>%
  filter(monday >= "2020-02-03") %>% # use this jawn to set "week 0" for the delta calculation
  mutate(delta = daily_ridership - first(daily_ridership)) %>% 
  mutate(delta_p = delta / first(daily_ridership))

# df that shows delta of ridership for each income tract over time
income_tract_linear <- tracts_weekly_ridership %>%
  group_by(income_bucket, monday) %>%
  summarise(daily_ridership = sum(ridership, na.rm = TRUE)) %>%
  arrange(monday) %>%
  filter(monday >= "2020-02-03" & income_bucket %>% is.na() == FALSE) %>% # use this jawn to set "week 0" for the delta calculation
  mutate(delta = daily_ridership - first(daily_ridership)) %>% 
  mutate(delta_p = delta / first(daily_ridership))

```

## Data Generation
See GitHub repository for documentation of methodology and approach. In short, data is taken from weekly samples of SEPTA bus APC data and is used to generate daily ridership at each stop. The data is not comprehensive and it is not well audited/cleaned, but it is the best snapshot of what has happened to transit ridership on a weekly basis through this crisis. 

### Data Sources:
Data was taken from: 

- SEPTA Automated Passenger Count Date + Turnstile Data (Not yet incorporated)
- ACS 2018 5-Year Estimates
- Misc. shapes pulled from opendataphilly

## Describing the Data
TODO - show some basic descriptive statistics of the data

## Visualizing the Data

```{r weekly_ridership, echo=FALSE}
# visualize that weekly ridership - yikes
ggplot(weekly_ridership, aes(x = as_date(monday), y = daily_ridership)) +
  geom_bar(position="dodge", stat="identity") + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 70000), breaks = seq(0, 70000, by=10000)) +
  scale_x_date(date_breaks = "1 week") + 
  xlab("Week Starting With") + 
  ylab("Estimated Average Daily Bus Ridership") + 
  ggtitle("SEPTA Daily Bus Ridership (2020)") + 
  theme_phl() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

#basic scatter plot of ridership and income - should scale points by total ridership?
ggplot(comp %>% filter(pct_change < 0.25), aes(x = med_income, y = pct_change)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  scale_y_continuous(labels = scales::percent) + 
  theme_phl()
```

### Examining Ridership by Census Tract

```{r tract_level, echo=FALSE, warning=FALSE}
# need to lable n() for each bar
p1 <- ggplot(income_tracts %>% filter(pct_change < 0.25), aes(x = income_bucket, y = pct_change)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent) + 
  scale_x_discrete(labels = c("Very Low Income (< 20k)", "Low Income (<35k)", "Med Income (<50k)", "High Income (<100k)", "Very High (>100k)")) +
  coord_cartesian(ylim = c(-1, -.5)) +
  theme_phl(base_size = 9) +
  #ggtitle("Bus Ridership Changes by Income Level of Census Tract") +
  ylab("Bus Ridership Change Post-COVID") + 
  xlab("Census Tracts Binned by Income Level") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95))

income_tracts_ccd <- comp %>%
  group_by(income_bucket, ccd.x) %>%
  summarise(pre_covid = sum(pre_covid_ridership),
            post_covid = sum(post_covid_ridership),
            diff = post_covid - pre_covid,
            pct_change = diff / pre_covid) %>%
  mutate(ccd = case_when(
    ccd.x == TRUE ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(is.na(income_bucket) == FALSE & ccd == FALSE)

p2 <- ggplot(income_tracts_ccd, aes(x = income_bucket, y = pct_change)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent) + 
  scale_x_discrete(labels = c("Very Low Income (< 20k)", "Low Income (<35k)", "Med Income (<50k)", "High Income   (<100k)", "Very High (>100k)")) + 
  coord_cartesian(ylim = c(-1, -.5)) +
  theme_phl(base_size = 9) +
  #ggtitle("Bus Ridership Changes by Income Level of Census Tract") +
  ylab("Bus Ridership Change Post-COVID") + 
  xlab("Census Tracts Binned by Income Level (Excluding Center City)") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95))

grid.arrange(p1, p2, nrow = 1,
             top = textGrob("Bus Ridership Changes by Income Level of Census Tract", 
                            gp=gpar(fontsize=10,fontfamily="Montserrat")))
```

``` {r income_tract_linear, echo=FALSE}
# label the theme better
ggplot(income_tract_linear, aes(x = monday, y = delta_p, color = income_bucket)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) + 
  xlab("Week Starting With") + 
  ylab("Daily Bus Ridership Change") + 
  ggtitle("") + 
  theme_phl()
```

``` {r planning_district, echo=FALSE} 
# visualize ridership change by planning district

ggplot(income_planning, aes(x = reorder(DIST_NAME.x, avg_med_income), y = pct_change, fill = avg_med_income)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent) + 
  coord_cartesian(ylim = c(-1, -.5)) + 
  theme_phl() + 
  scale_fill_phl(palette = "blues", discrete = F) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95)) +
  xlab("Planning District (Sorted by Income)") +
  ylab("Bus Ridership Change Post-COVID")

# need to put district names at the ends of the lines and vary the lines more to make it useful
ggplot(income_planning_linear, aes(x = as_date(monday), y = delta_p, color = DIST_NAME)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) + 
  scale_x_date(date_breaks = "1 week") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95)) +
  xlab("Week Starting With") + 
  ylab("Daily Bus Ridership Change") + 
  ggtitle("") + 
  theme_phl()
```

## Modeling Behavior
Goal here will be to look back and see what demographics were key predictors of a neighobrhood's ability to isolate and limit travel via transit. 

Things to look at:

- Income
- Race
- Car Ownership
- Education
- commute modes
- industry characteristics

Note: this assumes that the bus ridership of a neighborhood mirrors the demographic makeup of a neighborhood, which we know is not very true. Bus ridership in a neighobhrood is going to be composed of people with certain characteristics - e.g. work destinations that are transit accessible; lower incomes; and probably more.

ther path to look at: route level ridership loaded w/ surveyed rider demographics (pending data from SEPTA). 

### Correlations
```{r, echo=FALSE}
model_vars1 <- c("pct_change", "med_age", "med_income")
model_vars1_df <- comp[model_vars1]
model_vars1_df$geometry <- NULL
model_var1 <- cor(model_vars1_df, use = "pairwise.complete.obs")
corrplot(model_var1, method='number')

model_vars2 <- c("pct_change", "education_HS_p", "education_BA_p", "education_MS_p")
model_vars2_df <- comp[model_vars2]
model_vars2_df$geometry <- NULL
model_var2 <- cor(model_vars2_df, use = "pairwise.complete.obs")
corrplot(model_var2, method='number')

model_vars3 <- c("pct_change", "no_car_pct", "yes_car_pct")
model_vars3_df <- comp[model_vars3]
model_vars3_df$geometry <- NULL
model_var3 <- cor(model_vars3_df, use = "pairwise.complete.obs")
corrplot(model_var3, method='number')

model_vars4 <- c("pct_change", "white_pct", "black_pct", "asian_pct", "latinx_pct")
model_vars4_df <- comp[model_vars4]
model_vars4_df$geometry <- NULL
model_var4 <- cor(model_vars4_df, use = "pairwise.complete.obs")
corrplot(model_var4, method='number')

model_vars5 <- c("pct_change", "car_pct", "transit_pct", "bike_pct", "walk_pct", "wfh_pct")
model_vars5_df <- comp[model_vars5]
model_vars5_df$geometry <- NULL
model_var5 <- cor(model_vars5_df, use = "pairwise.complete.obs")
corrplot(model_var5, method='number')

model_vars6 <- c("pct_change", "industry1_pct", "industry2_pct", "industry3_pct", "industry4_pct", "industry5_pct", "industry6_pct", "industry7_pct",
                 "industry8_pct", "industry9_pct", "industry10_pct", "industry11_pct", "industry12_pct", "industry13_pct")
model_vars6_df <- comp[model_vars6]
model_vars6_df$geometry <- NULL
model_var6 <- cor(model_vars6_df, use = "pairwise.complete.obs")
corrplot(model_var6, method='number', number.cex=0.45)
```

```{r, echo = false}
#models with med_age, med_income, and % variables on education attainment, car ownership, and race
mod.a1 <- lm(pct_change ~ med_age + med_income + education_HS_p + education_BA_p + education_MS_p 
           + no_car_pct + yes_car_pct + white_pct + black_pct + asian_pct + latinx_pct, data = comp)
summary(mod.a1)


step(mod.a1, direction = "backward") #use stepwise backward function to get the best fiting model

mod.a2 <- lm(pct_change ~ med_age + med_income + education_HS_p + 
             education_MS_p + white_pct + black_pct + latinx_pct, data = comp)
summary(mod.a2)


#model with med_age, med_income, and % variables on education attainment, car ownership, and race, and commute
mod.b1 <- lm(pct_change ~ med_age + med_income + education_HS_p + education_BA_p + education_MS_p 
             + no_car_pct + yes_car_pct + white_pct + black_pct + asian_pct + latinx_pct 
             + car_pct + transit_pct + bike_pct + walk_pct + wfh_pct, data = comp)
summary(mod.b1)

step(mod.b1, direction = "backward") #use stepwise backward function to get the best fiting model

mod.b2 <- lm(pct_change ~ med_income 
                        + education_MS_p 
                        + no_car_pct 
                        + car_pct, 
                        data = comp)
summary(mod.b2)


#model with med_age, med_income, and % variables on education attainment, car ownership, and race, commute, and industry data

mod.c1 <- lm(pct_change ~ med_age + med_income + education_HS_p + education_BA_p + education_MS_p 
             + no_car_pct + yes_car_pct + white_pct + black_pct + asian_pct + latinx_pct + car_pct + transit_pct + bike_pct + walk_pct + wfh_pct 
             + industry1_pct + industry2_pct + industry3_pct + industry4_pct + industry5_pct + industry6_pct + industry7_pct
             + industry8_pct + industry9_pct + industry10_pct + industry11_pct + industry12_pct + industry13_pct, data = comp)
summary(mod.c1)

step(mod.c1, direction = "backward") #use stepwise backward function to get the best fiting model

mod.c2 <- lm(pct_change ~ med_age 
                        + med_income 
                        + education_MS_p 
                        + no_car_pct 
                        + car_pct 
                        + wfh_pct 
                        + industry2_pct 
                        + industry3_pct 
                        + industry5_pct 
                        + industry9_pct 
                        + industry10_pct 
                        + industry12_pct, 
                          data = comp)
summary(mod.c2)

## take out the non-significant variables from "mod_c2"
mod.c3 <- lm(pct_change ~ 
             + med_income 
             + education_MS_p 
             + no_car_pct 
             + car_pct 
             + industry5_pct 
             + industry9_pct 
             + industry10_pct 
             + industry12_pct, 
              data = comp)
summary(mod.c3)

#model with variables of interest

mod.d1<- lm(pct_change ~ med_income + education_BA_p + education_MS_p
             + yes_car_pct + white_pct + asian_pct + car_pct + transit_pct
             + industry1_pct + industry2_pct + industry3_pct + industry4_pct + industry5_pct + industry6_pct + industry7_pct
             + industry8_pct + industry9_pct + industry10_pct + industry11_pct + industry12_pct + industry13_pct, data = comp)
summary(mod.d1)

step(mod.d1, direction = "backward") #use stepwise backward function to get the best fiting model

mod.d2 <- lm(pct_change ~ med_income + education_MS_p + yes_car_pct + 
               car_pct + transit_pct + industry2_pct + industry3_pct + industry5_pct + 
               industry9_pct + industry10_pct + industry12_pct, data = comp)
summary(mod.d2)
```

```{r, results='asis', echo= FALSE}
stargazer(mod.a1, mod.a2, mod.b1, mod.b2, mod.c1, mod.c2, mod.c3, mod.d1, mod.d2, column.labels = c("mod.a1", "mod.a2", "mod.b1", "mod.b2", "mod.c1", "mod.c2", "mod.c3", "mod.d1", "mod.d2"), type ="html")
```
